---
title: "Cooperative Summary"
author: "Rachel Brooks"
format: html
editor: visual
---

## Import packages and clean-up data

```{r}

library(modelsummary)
library(tidyverse)
library(data.table)
library(ggplot2)
library(readxl)
library(dplyr)
library(googlesheets4)
library(webshot2)

coop <- read_sheet("https://docs.google.com/spreadsheets/d/1_ydIN0FOKdNB5z35uV0qqcNfAGHdt1HGheUmDc5B1LI/edit?gid=1569662549#gid=1569662549", sheet = "Merged Data")

# Clean-up weird 'Capture_Date' formatting
coop$Capture_Date <- sapply(coop$Capture_Date, function(x) {
  if (length(x) == 0) {
    NA_character_
  } else if (is.character(x)) {
    # Convert m/d/yyyy to ISO yyyy-mm-dd string
    as.character(as.Date(x, format = "%m/%d/%Y"))
  } else if (is.numeric(x)) {
    # Convert Excel serial to ISO yyyy-mm-dd string
    as.character(as.Date(x, origin = "1899-12-30"))
  } else {
    NA_character_
  }
})

# Filter out rows where there is no capture date 
# (filters out 15 samples as of 8/11/2025)
coop <- coop[!is.na(coop$Capture_Date), ]



```

## Assign vessels to CRFS districts and add species common names

```{r, warning = FALSE, message = FALSE, echo = FALSE}

unique_vessels <- unique(coop$Vessel)
print(unique_vessels)

coop <- coop %>%
  mutate(CRFS_District = case_when(

#District 1 - South
Vessel %in% c("Betty-O","New Del Mar","Eldorado","Amigo","(Newport) Patriot","Western Pride","Thunderbird","Fury","Sum Fun","Outer Limits","Premier" ) ~ "District 1",

# District 2 - Channel  
Vessel %in% c("Endeavor","Island Spirit","Coral Sea","Stardust","Mirage") ~ "District 2",

# District 3 - Central    
Vessel %in% c("Infinity","Legacy","Big Anchovy","Kahuna","Last Meal","The Marianne","Black Pearl","Fiesta","(Avila) Patriot") ~ "District 3",  

# District 4 - San Francisco    
Vessel %in% c("Dragonfly","Big Eye","New Sea Angler","Salty Lady","Lovely Martha","Sea Wolf","Queen of Hearts","New Captain Pete","Riptide") ~ "District 4",

# District 5 - Wine   
Vessel %in% c("Sea Hawk","Saylor Ann","Kyndall Lynn","Telstar") ~ "District 5",

# District 6 - Redwood    
Vessel %in% c("Onyx","Out Of The Box","New Reel Steel","Scrimshaw","Fishy Business") ~ "District 6",  # Changed from `==` to `%in%`
TRUE ~ "Unknown"  # Default for unmatched names
  ))

#Import PacFIN Species Table
PacFIN_Codes <- read_excel("PacFIN_Codes.xlsx")
coop <- merge(coop, PacFIN_Codes, by = "PacFIN_Species_Code", all.x = TRUE)  # Left join

coop <- coop %>%
  filter(!is.na(PacFIN_Species_Code))

coop$Species_Name <- recode_factor(coop$Species_Name,
                                      'Blue Rockfish' = "Blue/Deacon Rockfish",
                                      'Deacon Rockfish' = "Blue/Deacon Rockfish")

coop$Species_Name <- recode_factor(coop$Species_Name,
                                'Vermilion Rockfish' = "Vermilion/Sunset Rockfish")

#write.csv(Coop_Data, "coop-data.csv", row.names = FALSE)
```

## Annual summary of cooperative data by CRFS district

```{r}

library(dplyr)
library(lubridate)
library(tidyr)
library(gt)

# Build wide summary table
district_summary_wide <- coop %>%
  mutate(year = year(Capture_Date)) %>%
  filter(year %in% 2022:2025) %>%
  group_by(CRFS_District, year) %>%
  summarise(n_fish = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = year,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  mutate(Total = `2022` + `2023` + `2024` + `2025`)

# Reorder districts by number
district_summary_wide <- district_summary_wide %>%
  mutate(dist_num = as.numeric(gsub("\\D", "", CRFS_District))) %>%
  arrange(dist_num) %>%
  select(-dist_num)

# Add coastwide total row
district_summary_wide <- district_summary_wide %>%
  bind_rows(
    summarise(
      .,
      CRFS_District = "Coastwide Total",
      `2022` = sum(`2022`, na.rm = TRUE),
      `2023` = sum(`2023`, na.rm = TRUE),
      `2024` = sum(`2024`, na.rm = TRUE),
      `2025` = sum(`2025`, na.rm = TRUE),
      Total  = sum(Total,  na.rm = TRUE)
    )
  )

# Format for manuscript
district_table <- district_summary_wide %>%
  gt() %>%
  cols_label(
    CRFS_District = "District",
    `2022` = "2022",
    `2023` = "2023",
    `2024` = "2024",
    `2025` = "2025",
    Total  = "Total"
  ) %>%
  fmt_number(
    columns = c(`2022`, `2023`, `2024`, `2025`, Total),
    decimals = 0
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(columns = Total)
  ) %>%
  cols_width(
    CRFS_District ~ px(130),
    everything() ~ px(90)
  ) %>%
  tab_options(
    column_labels.padding = px(12),
    data_row.padding = px(12)
  ) %>%
  # Bold the Coastwide Total row
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(
      rows = CRFS_District == "Coastwide Total"
    )
  )

district_table

gtsave(district_table, "district_table.pdf")

```

## Annual summary of cooperative data by species 

```{r}

# Build wide summary table
species_summary_wide <- coop %>%
  mutate(year = year(Capture_Date)) %>%
  filter(year %in% 2022:2025) %>%
  filter(!is.na(Species_Name), Species_Name != "") %>%
  group_by(Species_Name, Scientific_Name, year) %>%       
  summarise(n_fish = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = year,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  select(Species_Name, Scientific_Name, `2022`, `2023`, `2024`, `2025`) %>%
  mutate(Total = `2022` + `2023` + `2024` + `2025`) %>%
  arrange(Scientific_Name)

# Format for manuscript
species_table <- species_summary_wide %>%
  gt() %>%
  cols_label(
    Scientific_Name = "Species",
    Species_Name = "Common Name",
    `2022` = "2022",
    `2023` = "2023",
    `2024` = "2024",
    `2025` = "2025",
    Total  = "Total"
  ) %>%
  fmt_number(
    columns = c(`2022`, `2023`, `2024`, `2025`, Total),
    decimals = 0
  ) %>%
  cols_align(
    align = "left",           # force left alignment
    columns = Species_Name
  ) %>%
     tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = "Scientific_Name")
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(columns = Total)
  ) %>%
  cols_width(
    Scientific_Name ~ px(300),
    Species_Name ~ px(200),
    everything() ~ px(90)
  ) %>%
  tab_options(
    column_labels.padding = px(12),
    data_row.padding = px(12)
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(
      rows = Species_Name == "Coastwide Total"
    )
  )

species_table

gtsave(species_table, "species_table.pdf")

```

## Summary of species by district

```{r}

# Define your ordered districts (adjust names exactly as in your data)
ordered_districts <- c(
  "District 1",
  "District 2",
  "District 3",
  "District 4",
  "District 5",
  "District 6"
)

species_by_district_wide <- coop %>%
  filter(!is.na(Species_Name), Species_Name != "") %>%
  group_by(Species_Name, Scientific_Name, CRFS_District) %>%
  summarise(n_fish = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = CRFS_District,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  # Reorder columns: first species columns, then districts in desired order, then Total
  mutate(Total = rowSums(across(all_of(ordered_districts)))) %>%
  select(
    Species_Name,
    Scientific_Name,
    all_of(ordered_districts),
    Total
  ) %>%
  arrange(Scientific_Name)

numeric_cols <- setdiff(colnames(species_by_district_wide), c("Species_Name", "Scientific_Name"))

species_by_district_table <- species_by_district_wide %>%
  gt() %>%
  cols_label(
    Species_Name = "Common Name",
    Scientific_Name = "Species"
  ) %>%
  fmt_number(
    columns = numeric_cols,
    decimals = 0
  ) %>%
  cols_align(
    align = "left",
    columns = c(Species_Name, Scientific_Name)
  ) %>%
   tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = "Scientific_Name")
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(columns = "Total")
  ) %>%
  cols_width(
    Scientific_Name ~ px(300),
    Species_Name ~ px(200),
    everything() ~ px(90)
  ) %>%
  tab_options(
    column_labels.padding = px(12),
    data_row.padding = px(12)
  )

species_by_district_table

gtsave(species_by_district_table, "species_by_district_table.pdf")

```

```{r, fig.width=15, fig.height=10, out.width="80%"}

# Calculate the abundance of each species
Abundance_Coastwide <- Coop_Data %>%
  group_by(Species_Name) %>%
  summarise(Abundance = n()) %>%
  arrange(desc(Abundance)) %>%
  top_n(12, Abundance)  # Get the top 10 most abundant species

# Filter the dataset to include only top 10 species
top_12_coastwide <- Coop_Data %>%
  filter(Species_Name %in% Abundance_Coastwide$Species_Name)

# Density plot
ggplot(top_12_coastwide, aes(x = Fork_Length_mm, fill = Sex)) +
 geom_histogram(aes(y = ..count..), bins = 20, alpha = 1, position = "identity", color = "black") +  # Histogram with density on y-axis
  geom_density(alpha = 0.6, color = "black") +  # Kernel density plot overlaid
  facet_wrap(~Species_Name) +  # Create separate panels for each species
  labs(title = "",
       x = "Fork Length (mm)",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("tomato", "dodgerblue", "grey")) +
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.title = element_text(size = 22),  # Adjust axis title size
    axis.text = element_text(size = 22),   # Adjust axis label text size
    plot.title = element_text(size = 22),  # Adjust plot title size
    strip.text = element_text(size = 22),   # Adjust facet label size
    legend.title = element_text(size = 22), # Increase legend title size
    legend.text = element_text(size = 22),  # Increase legend text size
    legend.key.size = unit(1, "cm")      # Increase the size of the legend keys
  )

```

## District 6 - Redwood

Includes F/V Onyx, F/V Out OF The Box, F/V New Reel Steel, F/V Scrimshaw

```{r, fig.width=12, fig.height=10, out.width="80%", fig.cap="Length Frequency Distribution of Top 6 Most Abundant Species in Distrct 6 - Redwood"}

D6 <- Coop_Data %>%
  filter(CRFS_District == "District 6 - Redwood")
datasummary_crosstab(Species_Name ~ Year, 
          statistic = 1 ~ N, 
          title = "Table 7: Summary of species sampled 2022-2024 for District 6 - Redwood", data = D6)

# Calculate the abundance of each species
Abundance_D6 <- D6 %>%
  filter(Year == "2024") %>%
  group_by(Species_Name) %>%
  summarise(Abundance = n()) %>%
  arrange(desc(Abundance)) %>%
  top_n(6, Abundance)  # Get the top 10 most abundant species

# Filter the dataset to include only top 10 species
top_6_species_D6 <- D6 %>%
  filter(Species_Name %in% Abundance_D6$Species_Name)

# Density plot
ggplot(top_6_species_D6, aes(x = Fork_Length_mm, fill = Sex)) +
geom_histogram(aes(y = ..count..), bins = 20, alpha = 1, position = "identity", color = "black") +  # Histogram with density on y-axis
  geom_density(alpha = 0.6, color = "black") +  # Kernel density plot overlaid
  facet_wrap(~Species_Name) +  # Create separate panels for each species
  labs(title = "",
       x = "Fork Length (mm)",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("tomato", "dodgerblue", "grey")) +
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.title = element_text(size = 22),  # Adjust axis title size
    axis.text = element_text(size = 22),   # Adjust axis label text size
    plot.title = element_text(size = 22),  # Adjust plot title size
    strip.text = element_text(size = 22),   # Adjust facet label size
    legend.title = element_text(size = 22), # Increase legend title size
    legend.text = element_text(size = 22),  # Increase legend text size
    legend.key.size = unit(1, "cm")      # Increase the size of the legend keys
  )

```

## District 5 - Wine

Includes F/V Kyndall Lynn, F/V Sea Hawk, F/V Saylor Ann

```{r, fig.width=12, fig.height=10, out.width="80%", fig.cap="Length Frequency Distribution of Top 6 Most Abundant Species in Distrct 5 - Redwood"}

D5 <- Coop_Data %>%
  filter(CRFS_District == "District 5 - Wine")
datasummary_crosstab(Species_Name ~ Year, 
          statistic = 1 ~ N, 
          title = "Table 7: Summary of species sampled 2022-2024 for District 6 - Redwood", data = D5)

# Calculate the abundance of each species
Abundance_D5 <- D5 %>%
  filter(Year == "2024") %>%
  group_by(Species_Name) %>%
  summarise(Abundance = n()) %>%
  arrange(desc(Abundance)) %>%
  top_n(6, Abundance)  # Get the top 10 most abundant species

# Filter the dataset to include only top 10 species
top_6_species_D5 <- D5 %>%
  filter(Species_Name %in% Abundance_D5$Species_Name)

# Density plot
ggplot(top_6_species_D5, aes(x = Fork_Length_mm, fill = Sex)) +
geom_histogram(aes(y = ..count..), bins = 20, alpha = 1, position = "identity", color = "black") +  # Histogram with density on y-axis
  geom_density(alpha = 0.6, color = "black") +  # Kernel density plot overlaid
  facet_wrap(~Species_Name) +  # Create separate panels for each species
  labs(title = "",
       x = "Fork Length (mm)",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("tomato", "dodgerblue", "grey")) +
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.title = element_text(size = 22),  # Adjust axis title size
    axis.text = element_text(size = 22),   # Adjust axis label text size
    plot.title = element_text(size = 22),  # Adjust plot title size
    strip.text = element_text(size = 22),   # Adjust facet label size
    legend.title = element_text(size = 22), # Increase legend title size
    legend.text = element_text(size = 22),  # Increase legend text size
    legend.key.size = unit(1, "cm")      # Increase the size of the legend keys
  )


```

## District 4 - San Francisco

```{r, fig.width=12, fig.height=10, out.width="80%", fig.cap="Length Frequency Distribution of Top 6 Most Abundant Species in Distrct 5 - Redwood"}

D4 <- Coop_Data %>%
  filter(CRFS_District == "District 4 - San Francisco")
datasummary_crosstab(Species_Name ~ Year, 
          statistic = 1 ~ N, 
          title = "Table 5: Summary of species sampled 2022-2024 for District 4 - San Francisco", data = D4)

# Calculate the abundance of each species
Abundance_D4 <- D4 %>%  
  filter(Year == "2024") %>%
  group_by(Species_Name) %>%
  summarise(Abundance = n()) %>%
  arrange(desc(Abundance)) %>%
  top_n(6, Abundance)  # Get the top 10 most abundant species

# Filter the dataset to include only top 10 species
top_6_species_D4 <- D4 %>%
  filter(Species_Name %in% Abundance_D4$Species_Name)

# Density plot
ggplot(top_6_species_D4, aes(x = Fork_Length_mm, fill = Sex)) +
geom_histogram(aes(y = ..count..), bins = 20, alpha = 1, position = "identity", color = "black") +  # Histogram with density on y-axis
  geom_density(alpha = 0.6, color = "black") +  # Kernel density plot overlaid
  facet_wrap(~Species_Name) +  # Create separate panels for each species
  labs(title = "",
       x = "Fork Length (mm)",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("tomato", "dodgerblue", "grey")) +
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.title = element_text(size = 22),  # Adjust axis title size
    axis.text = element_text(size = 22),   # Adjust axis label text size
    plot.title = element_text(size = 22),  # Adjust plot title size
    strip.text = element_text(size = 22),   # Adjust facet label size
    legend.title = element_text(size = 22), # Increase legend title size
    legend.text = element_text(size = 22),  # Increase legend text size
    legend.key.size = unit(1, "cm")      # Increase the size of the legend keys
  )

```

## District 3 - Central

```{r, fig.width=12, fig.height=10, out.width="80%", fig.cap="Length Frequency Distribution of Top 6 Most Abundant Species in Distrct 5 - Redwood"}

D3 <- Coop_Data %>%
  filter(CRFS_District == "District 3 - Central")
datasummary_crosstab(Species_Name ~ Year, 
          statistic = 1 ~ N, 
          title = "Table 4: Summary of species sampled 2022-2024 for District 3 - Central", data = D3)

# Calculate the abundance of each species
Abundance_D3 <- D3 %>%  
  filter(Year == "2024") %>%
  group_by(Species_Name) %>%
  summarise(Abundance = n()) %>%
  arrange(desc(Abundance)) %>%
  top_n(6, Abundance)  # Get the top 10 most abundant species

# Filter the dataset to include only top 10 species
top_6_species_D3 <- D3 %>%
  filter(Species_Name %in% Abundance_D3$Species_Name)

# Density plot
ggplot(top_6_species_D3, aes(x = Fork_Length_mm, fill = Sex)) +
geom_histogram(aes(y = ..count..), bins = 20, alpha = 1, position = "identity", color = "black") +  # Histogram with density on y-axis
  geom_density(alpha = 0.6, color = "black") +  # Kernel density plot overlaid
  facet_wrap(~Species_Name) +  # Create separate panels for each species
  labs(title = "",
       x = "Fork Length (mm)",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("tomato", "dodgerblue", "grey")) +
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.title = element_text(size = 22),  # Adjust axis title size
    axis.text = element_text(size = 22),   # Adjust axis label text size
    plot.title = element_text(size = 22),  # Adjust plot title size
    strip.text = element_text(size = 22),   # Adjust facet label size
    legend.title = element_text(size = 22), # Increase legend title size
    legend.text = element_text(size = 22),  # Increase legend text size
    legend.key.size = unit(1, "cm")      # Increase the size of the legend keys
  )



```

## District 2 - Channel

```{r, fig.width=12, fig.height=10, out.width="80%", fig.cap="Length Frequency Distribution of Top 6 Most Abundant Species in Distrct 5 - Redwood"}

D2 <- Coop_Data %>%
  filter(CRFS_District == "District 2 - Channel")
datasummary_crosstab(Species_Name ~ Year, 
          statistic = 1 ~ N, 
          title = "Table 3: Summary of species sampled 2022-2024 for District 2 - Channel", data = D2)


# Calculate the abundance of each species
Abundance_D2 <- D2 %>%  
  filter(Year == "2024") %>%
  group_by(Species_Name) %>%
  summarise(Abundance = n()) %>%
  arrange(desc(Abundance)) %>%
  top_n(6, Abundance)  # Get the top 10 most abundant species

# Filter the dataset to include only top 10 species
top_6_species_D2 <- D2 %>%
  filter(Species_Name %in% Abundance_D2$Species_Name)

# Density plot
ggplot(top_6_species_D2, aes(x = Fork_Length_mm, fill = Sex)) +
geom_histogram(aes(y = ..count..), bins = 20, alpha = 1, position = "identity", color = "black") +  # Histogram with density on y-axis
  geom_density(alpha = 0.6, color = "black") +  # Kernel density plot overlaid
  facet_wrap(~Species_Name) +  # Create separate panels for each species
  labs(title = "",
       x = "Fork Length (mm)",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("tomato", "dodgerblue", "grey")) +
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.title = element_text(size = 22),  # Adjust axis title size
    axis.text = element_text(size = 22),   # Adjust axis label text size
    plot.title = element_text(size = 22),  # Adjust plot title size
    strip.text = element_text(size = 22),   # Adjust facet label size
    legend.title = element_text(size = 22), # Increase legend title size
    legend.text = element_text(size = 22),  # Increase legend text size
    legend.key.size = unit(1, "cm")      # Increase the size of the legend keys
  )



```

## District 1 - South

```{r, fig.width=12, fig.height=10, out.width="80%", fig.cap="Length Frequency Distribution of Top 6 Most Abundant Species in District 1 - South"}

D1 <- Coop_Data %>%
  filter(CRFS_District == "District 1 - South")
datasummary_crosstab(Species_Name ~ Year, 
          statistic = 1 ~ N, 
          title = "Table 2: Summary of species sampled 2022-2024 for District 1 - South", data = D1)

# Calculate the abundance of each species
Abundance_D1 <- D1 %>%  
  filter(Year == "2024") %>%
  group_by(Species_Name) %>%
  summarise(Abundance = n()) %>%
  arrange(desc(Abundance)) %>%
  top_n(6, Abundance)  # Get the top 10 most abundant species

# Filter the dataset to include only top 10 species
top_6_species_D1 <- D1 %>%
  filter(Species_Name %in% Abundance_D1$Species_Name)

# Density plot
ggplot(top_6_species_D1, aes(x = Fork_Length_mm, fill = Sex)) +
 geom_histogram(aes(y = ..count..), bins = 20, alpha = 1, position = "identity", color = "black") +  # Histogram with density on y-axis
  geom_density(alpha = 0.6, color = "black") +  # Kernel density plot overlaid
  facet_wrap(~Species_Name) +  # Create separate panels for each species
  labs(title = "",
       x = "Fork Length (mm)",
       y = "Count") +
  theme_minimal() +
  scale_fill_manual(values = c("tomato", "dodgerblue", "grey")) +
  theme(
    panel.grid = element_blank(),  # Remove grid lines
    axis.title = element_text(size = 22),  # Adjust axis title size
    axis.text = element_text(size = 22),   # Adjust axis label text size
    plot.title = element_text(size = 22),  # Adjust plot title size
    strip.text = element_text(size = 22),   # Adjust facet label size
    legend.title = element_text(size = 22), # Increase legend title size
    legend.text = element_text(size = 22),  # Increase legend text size
    legend.key.size = unit(1, "cm")      # Increase the size of the legend ke
```
