---
title: "Cooperative Summary"
author: "Rachel Brooks"
format: html
editor: visual
---

```{r, warning = FALSE, message = FALSE, echo = FALSE}

library(modelsummary)
library(tidyverse)
library(data.table)
library(ggplot2)
library(readxl)
library(dplyr)
library(googlesheets4)
library(webshot2)
library(tigris)
library(sf)

```

## Project Map

Map differentiating CRFS districts that our sampling scheme follows

```{r}

# Download California counties (simplified)
counties <- counties(state = "CA", cb = TRUE, class = "sf")

# Map counties to CRFS districts
district_map <- list(
  "District 1" = c("Los Angeles", "Orange", "San Diego"),
  "District 2" = c("Santa Barbara", "Ventura"),
  "District 3" = c("Santa Cruz", "Monterey", "San Luis Obispo"),
  "District 4" = c("Sonoma", "Marin", "San Francisco", "San Mateo",
                      "Alameda", "Contra Costa", "Solano", "Santa Clara"),
  "District 5" = c("Mendocino", "Humboldt"),
  "District 6" = c("Del Norte", "Humboldt")
)

counties$district <- NA
for (d in names(district_map)) {
  counties$district[counties$NAME %in% district_map[[d]]] <- d
}

# Custom colors
library(viridis)

district_colors <- viridis(n = 6)  # "M" = magma
names(district_colors) <- paste0("District ", 6:1)

# Compute centroids for labels
district_centroids <- counties %>%
  filter(!is.na(district)) %>%
  group_by(district) %>%
  summarise(geometry = st_union(geometry)) %>%
  st_centroid() %>%
  st_coordinates() %>%
  as.data.frame() %>%
  mutate(district = c("District 1", "District 2", "District 3", "District 4", "District 5", "District 6"))

# Manually adjust label positions
district_centroids <- district_centroids %>%
  mutate(
    X = case_when(
      district == "District 3" ~ X - 2,
      district == "District 2" ~ X - 2.5,
      district == "District 6" ~ X - 2,
      district == "District 4" ~ X - 2.5,
      district == "District 1" ~ X - 3.5,
      district == "District 5" ~ X - 2,
      TRUE ~ X
    ),
    Y = case_when(
      district == "District 3" ~ Y - 0.1,
      district == "District 2" ~ Y - 0.5,
      district == "District 6" ~ Y + 0.05,
      district == "District 4" ~ Y - 0.4,
      district == "District 1" ~ Y - 0.5,
      district == "District 5" ~ Y - 0.25,
      TRUE ~ Y
    )
  )

# Create California outline
ca_outline <- counties %>% st_union() %>% st_sf()

# Plot
ggplot() +
  # Background counties
  geom_sf(data = counties, fill = "grey80", color = "grey80") +
  # CRFS districts
  geom_sf(data = counties[!is.na(counties$district), ], 
          aes(fill = district), color = "black") +
  # California outline
  geom_sf(data = ca_outline, fill = NA, color = "black", size = 1) +
  # District labels
  geom_text(data = district_centroids, 
            aes(x = X, y = Y, label = district),
            color = "black", size = 4, fontface = "bold") +
  scale_fill_manual(values = district_colors) +
  coord_sf(xlim = c(-127, -114), ylim = c(32, 42)) +
  theme_minimal() +  # keep axes, grid lines, and ticks
  theme(
     panel.grid = element_blank(),   # remove gridlines
    axis.ticks = element_blank(),   # remove ticks
    axis.text  = element_blank(),   # remove tick labels
    axis.title = element_blank(),   # remove axis titles
    legend.position = "none",
    plot.title = element_text(size = 16, face = "bold")
  ) +
  labs(title = "")

```

## Import and clean-up data

1.  Pulls most up-to-date data from google sheets,

2.  Assigns CRFS districts to catch based on sampling port

3.  Appends PacFIN species information to data

4.  Renames blue and deacon rockfish as blue/deacon rockfish complex

5.  Renames vermilion rockfish as vermilion/sunset rockfish complex

```{r, warning = FALSE, message = FALSE, echo = FALSE}

coop <- read_sheet("https://docs.google.com/spreadsheets/d/1_ydIN0FOKdNB5z35uV0qqcNfAGHdt1HGheUmDc5B1LI/edit?gid=1569662549#gid=1569662549", sheet = "Merged Data")

# Clean-up weird 'Capture_Date' formatting
coop$Capture_Date <- sapply(coop$Capture_Date, function(x) {
  if (length(x) == 0) {
    NA_character_
  } else if (is.character(x)) {
    # Convert m/d/yyyy to ISO yyyy-mm-dd string
    as.character(as.Date(x, format = "%m/%d/%Y"))
  } else if (is.numeric(x)) {
    # Convert Excel serial to ISO yyyy-mm-dd string
    as.character(as.Date(x, origin = "1899-12-30"))
  } else {
    NA_character_
  }
})

# Filter out rows where there is no capture date 
# (filters out 15 samples as of 8/11/2025)
#coop <- coop[!is.na(coop$Capture_Date), ]

unique_vessels <- unique(coop$Vessel)
print(unique_vessels)

coop <- coop %>%
  mutate(CRFS_District = case_when(

#District 1 - South
Vessel %in% c("Betty-O","New Del Mar","Eldorado","Amigo","(Newport) Patriot","Western Pride","Thunderbird","Fury","Sum Fun","Outer Limits","Premier", "Malihini" ) ~ "District 1",

# District 2 - Channel  
Vessel %in% c("Endeavor","Island Spirit","Coral Sea","Stardust","Mirage") ~ "District 2",

# District 3 - Central    
Vessel %in% c("Infinity","Legacy","Big Anchovy","Kahuna","Last Meal","The Marianne","Black Pearl","Fiesta","(Avila) Patriot") ~ "District 3",  

# District 4 - San Francisco    
Vessel %in% c("Dragonfly","Big Eye","Sorta Salty","New Sea Angler","Salty Lady","Lovely Martha","Sea Wolf","Queen of Hearts","New Captain Pete","Riptide") ~ "District 4",

# District 5 - Wine   
Vessel %in% c("Sea Hawk","Saylor Ann","Kyndall Lynn","Telstar") ~ "District 5",

# District 6 - Redwood    
Vessel %in% c("Onyx","Out Of The Box","New Reel Steel","Scrimshaw","Fishy Business") ~ "District 6",  # Changed from `==` to `%in%`
TRUE ~ "Unknown"  # Default for unmatched names
  ))

#Import PacFIN Species Table
PacFIN_Codes <- read_excel("PacFIN_Codes.xlsx")
coop <- merge(coop, PacFIN_Codes, by = "PacFIN_Species_Code", all.x = TRUE)  # Left join

coop <- coop %>%
  filter(!is.na(PacFIN_Species_Code))

coop$Species_Name <- recode_factor(coop$Species_Name,
                                      'Blue Rockfish' = "Blue/Deacon Rockfish",
                                      'Deacon Rockfish' = "Blue/Deacon Rockfish")
coop$Scientific_Name <- recode_factor(coop$Scientific_Name,
                                      'Sebastes mystinus' = "Sebastes mystinus/diaconus",
                                      'Sebastes diaconus' = "Sebastes mystinus/diaconus")

coop$Species_Name <- recode_factor(coop$Species_Name,
                                'Vermilion Rockfish' = "Vermilion/Sunset Rockfish")

#write.csv(coop, "coop-data.csv", row.names = FALSE)

```

## Annual summary of cooperative data by CRFS district

```{r}

library(lubridate)
library(tidyr)
library(gt)

# Build wide summary table
district_summary_wide <- coop %>%
  group_by(CRFS_District, Year) %>%
  summarise(n_fish = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = Year,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  mutate(
    Total = rowSums(across(where(is.numeric)))
  )

# Reorder districts by number
district_summary_wide <- district_summary_wide %>%
  mutate(dist_num = as.numeric(gsub("\\D", "", CRFS_District))) %>%
  arrange(dist_num) %>%
  select(-dist_num)

# Add coastwide total row
district_summary_wide <- district_summary_wide %>%
  bind_rows(
    summarise(
      .,
      CRFS_District = "Coastwide Total",
      `2022` = sum(`2022`, na.rm = TRUE),
      `2023` = sum(`2023`, na.rm = TRUE),
      `2024` = sum(`2024`, na.rm = TRUE),
      `2025` = sum(`2025`, na.rm = TRUE),
      Total  = sum(Total,  na.rm = TRUE)
    )
  )

# Format for manuscript
district_table <- district_summary_wide %>%
  gt() %>%
  cols_label(
    CRFS_District = "District",
    `2022` = "2022",
    `2023` = "2023",
    `2024` = "2024",
    `2025` = "2025",
    Total  = "Total"
  ) %>%
  fmt_number(
    columns = c(`2022`, `2023`, `2024`, `2025`, Total),
    decimals = 0
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(columns = Total)
  ) %>%
  cols_width(
    CRFS_District ~ px(130),
    everything() ~ px(90)
  ) %>%
  tab_options(
    column_labels.padding = px(12),
    data_row.padding = px(12)
  ) %>%
  # Bold the Coastwide Total row
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(
      rows = CRFS_District == "Coastwide Total"
    )
  )

district_table

gtsave(district_table, "district_table.png")

write.csv(
  district_table,
  "district_table.csv",
  row.names = FALSE
)

```

## Annual summary of cooperative data by species

```{r}


# Build wide summary table
species_summary_wide <- coop %>%
  filter(Year %in% 2022:2025) %>%
  filter(!is.na(Species_Name), Species_Name != "") %>%
  group_by(Species_Name, Scientific_Name, Year) %>%
  summarise(n_fish = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = Year,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  mutate(
    Total = rowSums(across(where(is.numeric)))
  ) %>%
  bind_rows(
    summarise(
      .,
      Species_Name = "Coastwide Total",
      Scientific_Name = "",
      across(where(is.numeric), ~ sum(.x, na.rm = TRUE))
    )
  ) %>%
  mutate(
    is_total = Species_Name == "Coastwide Total"
  ) %>%
  arrange(
    is_total,                 # Coastwide Total last
    desc(Total),              # Most common species first
    Scientific_Name            # Alphabetical tie-breaker
  ) %>%
  select(-is_total)


# Format table for manuscript
species_table <- species_summary_wide %>%
  gt() %>%
  cols_label(
    Scientific_Name = "Species",
    Species_Name = "Common Name",
    `2022` = "2022",
    `2023` = "2023",
    `2024` = "2024",
    `2025` = "2025",
    Total  = "Total"
  ) %>%
  fmt_number(
    columns = c(`2022`, `2023`, `2024`, `2025`, Total),
    decimals = 0
  ) %>%
  cols_align(
    align = "left",
    columns = Species_Name
  ) %>%
  tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = Scientific_Name)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Total)
  ) %>%
  cols_width(
    Scientific_Name ~ px(300),
    Species_Name ~ px(200),
    everything() ~ px(90)
  ) %>%
  tab_options(
    column_labels.padding = px(12),
    data_row.padding = px(12)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = Species_Name == "Coastwide Total"
    )
  )

# Save table
gtsave(species_table, "species_table.png")

write.csv(
  species_summary_wide,
  "species_table.csv",
  row.names = FALSE
)

```

## Summary of species by district

```{r}

# Define your ordered districts (adjust names exactly as in your data)
ordered_districts <- c(
  "District 1",
  "District 2",
  "District 3",
  "District 4",
  "District 5",
  "District 6"
)

species_by_district_wide <- coop %>%
  filter(!is.na(Species_Name), Species_Name != "") %>%
  group_by(Species_Name, Scientific_Name, CRFS_District) %>%
  summarise(n_fish = n(), .groups = "drop") %>%
  pivot_wider(
    names_from = CRFS_District,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  # Reorder columns: first species columns, then districts in desired order, then Total
  mutate(Total = rowSums(across(all_of(ordered_districts)))) %>%
  select(
    Species_Name,
    Scientific_Name,
    all_of(ordered_districts),
    Total
  ) %>%
  arrange(Scientific_Name)

numeric_cols <- setdiff(colnames(species_by_district_wide), c("Species_Name", "Scientific_Name"))

species_by_district_table <- species_by_district_wide %>%
  gt() %>%
  cols_label(
    Species_Name = "Common Name",
    Scientific_Name = "Species"
  ) %>%
  fmt_number(
    columns = numeric_cols,
    decimals = 0
  ) %>%
  cols_align(
    align = "left",
    columns = c(Species_Name, Scientific_Name)
  ) %>%
   tab_style(
    style = cell_text(style = "italic"),
    locations = cells_body(columns = "Scientific_Name")
  ) %>%
  tab_style(
    style = list(cell_text(weight = "bold")),
    locations = cells_body(columns = "Total")
  ) %>%
  cols_width(
    Scientific_Name ~ px(300),
    Species_Name ~ px(200),
    everything() ~ px(90)
  ) %>%
  tab_options(
    column_labels.padding = px(12),
    data_row.padding = px(12)
  )

species_by_district_table

#gtsave(species_by_district_table, "species_by_district_table.png")



write.csv(
  species_by_district_wide,
  "species_district_table.csv",
  row.names = FALSE
)

```

## Summary of species by district by year (Melissa)

```{r}

# Define districts and years
ordered_districts <- c(
  "District 1", "District 2", "District 3",
  "District 4", "District 5", "District 6"
)
years <- 2022:2025

# Build the wide summary
species_summary_wide <- coop %>%
  filter(!is.na(Species_Name), Species_Name != "") %>%
  filter(Year %in% years) %>%
  group_by(Species_Name, Scientific_Name, CRFS_District, Year) %>%
  summarise(n_fish = n(), .groups = "drop") %>%
  # Combine district and year into one column for pivoting
  mutate(District_Year = paste(CRFS_District, Year, sep = "_")) %>%
  select(-CRFS_District, -Year) %>%
  pivot_wider(
    names_from = District_Year,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  # Optional: reorder columns so districts × years are in consistent order
  {
    district_year_cols <- as.vector(outer(ordered_districts, years, paste, sep = "_"))
    select(., Species_Name, Scientific_Name, any_of(district_year_cols))
  } %>%
  # Add a total column per species
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(desc(Total), Scientific_Name)

write.csv(
  species_summary_wide,
  "species_summary_one_row_per_species.csv",
  row.names = FALSE
)
```

## Summary of species by port by year (Melissa)

```{r}

# Define ports and years
ordered_ports <- c(
"Crescent City","Eureka","Shelter Cove","Fort Bragg","Bodega Bay","San Francisco","Sausalito","Emeryville","Pillar Point","Santa Cruz","Moss Landing","Morro Bay","Avila Beach","Santa Barbara","Ventura","Channel Islands","Marina Del Rey","Long Beach","San Pedro","Newport","Dana Point","Mission Bay","Seaforth","San Diego","UNKNOWN"
)
years <- 2022:2025

# Build the wide summary by port
species_summary_by_port <- coop %>%
  filter(!is.na(Species_Name), Species_Name != "") %>%
  filter(Year %in% years) %>%
  group_by(Species_Name, Scientific_Name, Port, Year) %>%
  summarise(n_fish = n(), .groups = "drop") %>%
  # Combine port and year into one column for pivoting
  mutate(Port_Year = paste(Port, Year, sep = "_")) %>%
  select(-Port, -Year) %>%
  pivot_wider(
    names_from = Port_Year,
    values_from = n_fish,
    values_fill = 0
  ) %>%
  # Reorder columns so ports × years are in consistent order
  {
    port_year_cols <- as.vector(outer(ordered_ports, years, paste, sep = "_"))
    select(., Species_Name, Scientific_Name, any_of(port_year_cols))
  } %>%
  # Add a total column per species
  mutate(Total = rowSums(across(where(is.numeric)))) %>%
  arrange(desc(Total), Scientific_Name)

# Save as CSV
write.csv(
  species_summary_by_port,
  "species_summary_by_port_one_row_per_species.csv",
  row.names = FALSE
)
```

## Stacked bar plot

```{r}

#Two Copper Rockfish with UNKNOWN District - Assigned to Distrcit 2 because either Ventura or SB fish
coop_fixed <- coop %>%
  mutate(
    CRFS_District = if_else(
      Species_Name == "Copper Rockfish" & CRFS_District == "Unknown",
      "District 2",   
      CRFS_District
    )
  )
species_counts <- coop_fixed %>%
  filter(
    !is.na(Species_Name),
    Species_Name != "",
    !is.na(CRFS_District),
    CRFS_District != "UNKNOWN"
  ) %>%
  group_by(Species_Name, CRFS_District) %>%
  summarise(n = n(), .groups = "drop")

# Identify top 20 species
top_species <- species_counts %>%
  group_by(Species_Name) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(desc(total_n)) %>%
  slice_head(n = 20) %>%
  pull(Species_Name)

# Recode other species as "Other Groundfish"
species_counts <- species_counts %>%
  mutate(
    Species_Name = if_else(
      Species_Name %in% top_species,
      Species_Name,
      "Other Groundfish"
    )
  ) %>%
  group_by(Species_Name, CRFS_District) %>%
  summarise(n = sum(n), .groups = "drop")


# Re-aggregate after combining "Other Groundfish"
species_order <- species_counts %>%
  filter(Species_Name != "Other Groundfish") %>%
  group_by(Species_Name) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(desc(total_n)) %>%
  pull(Species_Name)

species_counts <- species_counts %>%
  mutate(
    Species_Name = factor(
      Species_Name,
      levels = c(species_order, "Other Groundfish")
    ),
    CRFS_District = fct_drop(factor(CRFS_District))
  )


# Plot
ggplot(species_counts,
       aes(x = Species_Name, y = n, fill = CRFS_District)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d(option = "D", direction = -1, drop = TRUE) +
  labs(
    x = "Species",
    y = "Count",
    fill = "CRFS District"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),   
    axis.ticks = element_blank(), 
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 5, r = 5, b = 5, l = 27)
  )

```

```{r}

library(forcats)

# 1. Identify top 5 species
top5_species <- coop %>%
  mutate(Fork_Length_mm = suppressWarnings(as.numeric(as.character(Fork_Length_mm)))) %>%
  filter(!is.na(Fork_Length_mm)) %>%
  count(Species_Name, sort = TRUE) %>%
  slice_max(n, n = 5) %>%
  pull(Species_Name)

# 2. Prepare cleaned plot data
plot_data <- coop %>%
  mutate(
    # Convert lengths to numeric
    Fork_Length_mm = suppressWarnings(as.numeric(as.character(Fork_Length_mm))),
    # Convert district to factor with correct order
    CRFS_District = factor(CRFS_District,
                           levels = c("District 6", "District 5", "District 4",
                                      "District 3", "District 2", "District 1")),
    # Factor for sex
    Sex = factor(Sex, levels = c("M", "F", "U"))
  ) %>%
  filter(Species_Name %in% top5_species) %>%
  filter(!is.na(Fork_Length_mm)) %>%
  # Fill missing district/sex combinations per species (keeps empty facets)
  complete(Species_Name, CRFS_District, Sex, fill = list(Fork_Length_mm = NA))

# Remove rows with missing district
plot_data <- plot_data %>%
  filter(!is.na(CRFS_District))

# 3. Function to make one plot per species
make_plot <- function(species_name) {
  ggplot(
    filter(plot_data, Species_Name == species_name, !is.na(Fork_Length_mm)),  # <- filter NA
    aes(x = Fork_Length_mm, fill = Sex)
  ) +
    geom_histogram(
      binwidth = 10, 
      color = "black", 
      position = "identity",
      alpha = 0.5) +
    facet_wrap(~ CRFS_District, ncol = 1, drop = FALSE) +
    scale_fill_manual(values = c("M" = "dodgerblue", "F" = "tomato", "U" = "gray60")) +
    labs(title = species_name,
         x = "Fork length (mm)",
         y = "Count",
         fill = "Sex") +
    theme_minimal(base_size = 14)+
    theme(panel.grid = element_blank())
}

# 4. Create plots for top 5 species
plots_list <- lapply(top5_species, make_plot)

# 5. Example: display first plot
plots_list[[1]]
plots_list[[2]]
plots_list[[3]]
plots_list[[4]]
plots_list[[5]]

pdf("Top5_Species_Length_Freq.pdf", width = 6, height = 12)  # increase height
lapply(plots_list, print)
dev.off()

for (i in seq_along(plots_list)) {
  species_name <- top5_species[i]
  safe_name <- gsub("[^[:alnum:]]+", "_", species_name)  # replace non-alphanumeric with _
  
  ggsave(
    filename = paste0(safe_name, "_ForkLength.jpg"),
    plot = plots_list[[i]],
    device = "jpeg",
    width = 6,
    height = 10,
    units = "in",
    dpi = 300
  )
}

```
